Structure Learning for $\VAR(p)$ Models

### 2019-05-01 paper.tex
Organizing and finishing / cleaning up some other proofs.

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{pa(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node
\newcommand{\ancn}[2]{\mathcal{A}_{#1}(#2)}  % nth ancestors of a node
\newcommand{\gpn}[2]{gp_{#1}(#2)}  % nth generation grandparents
\newcommand{\gcgpath}[4]{#1 \rightarrow #2_1 \rightarrow \cdots \rightarrow #2_{#3} \rightarrow #4}
\newcommand{\wtalpha}[2]{\widetilde{\alpha}(#1, #2)}  % Some notation for lem:pwgc_anc
\newcommand{\dist}[2]{\mathsf{d}(#1, #2)}  % Distance between things

@@

@@

  Moreover, if $j \not\in \anc{i}$ then $\exists u \in \anc{i} \cap \anc{j}$ such that $i$ is not contained in at least one $u \rightarrow \cdots \rightarrow j$ path.
\end{proposition}

@@

It seems intuitive that a converse of \ref{prop:ancestor_properties}
would hold, i.e. $j \in \anc{i} \implies j \pwgc i$.  Unfortunately,
this is not the case in general, as different paths through $\gcg$ can
lead to cancellation (see example \ref{ex:diamond_cancellation}).  In
fact, we do not even have $j \in \pa{i} \implies j \pwgc i$ (see
example \ref{ex:lag_cancellation}).

@@

The previous examples seem rather pathological since we have simply
constructed a case where different paths in the graph cancel exactly.
It may be possible to rid ourselves of these pathologies for instance
by defining a notion of ``robust'' pairwise causality where
``$j \pwgc _R i$'' if $j \pwgc i$ almost surely for a ``small'' random
perturbation of the system matrix.  However, we do not believe such a
modification of our theory can have useful practical implications
since the difference between paths that ``nearly cancel'' rather than
cancel exactly would in practice require inordinate amounts of data to
resolve.

@@

Examples of strongly causal graphs include directed trees (or
forests), and DAGs where each node has at most one parent.

\begin{lemma}
  \label{lem:still_strongly_causal}
  If $\gcg$ is a strongly causal graph, any subgraph formed by eliminating nodes
  and all of the in and out edges of the eliminated node is still strongly causal.
\end{lemma}
\begin{proof}
  If $\gcg$ has at most one path between any two nodes, there can only
  be fewer paths after removing nodes from $\gcg$.
\end{proof}

\begin{lemma}
  \label{prop:sc_graph_common_anc}
  In a strongly causal graph if $j \in \anc{i}$ and $k \in \anc{i} \cap \anc{j}$ then the unique path from $k$ to $i$ contains $j$.
\end{lemma}
\begin{proof}
  Suppose that there is a path from $k$ to $i$ which does not contain $j$.  In this case, there are multiple paths from $k$ to $i$ (one of which \textit{does} go through $j$ since $j \in \anc{i}$) which contradicts the assumption of strong causality.
\end{proof}

@@

\begin{proposition}
  If $\gcg$ is a strongly causal DAG and $i \pwgc j$ and $j \pwgc i$, then $i \not\in \pa{j}$ and $j \not\in \pa{i}$.  That is, a pairwise bidirectional edge indicates the absense of any edge in $\gcg$.
\end{proposition}
\begin{proof}
  By way of contradiction, suppose that $i \in \pa{j}$ and note that the case $j \in \pa{i}$ is symmetric.  We consider the two possibilities allowed by proposition \ref{prop:ancestor_properties} for $j \pwgc i$:  Firstly $j \in \anc{i}$ is impossible since $\gcg$ is assumed to be acyclic; \hl{Secondly} $u \in \anc{i} \cap \anc{j}$ \hl{leads to a contradiction} since there must be multiple paths, namely $u \rightarrow \cdots \rightarrow i \rightarrow \cdots \rightarrow j$ and $u \rightarrow \cdots \rightarrow j$ in $\gcg$, but this isn't possible since $\gcg$ is strongly causal, therefore we cannot have $i \in \pa{j}$. \hl{This requires a strengthening of proposition 3!}
\end{proof}

@@

  Firstly, we will establish a representation of $x_i(t)$ that involves $x_j(t)$.  Denote by $a_{r + 1} \rightarrow a_r \rightarrow \cdots \rightarrow a_1 \rightarrow a_0$ with $a_{r + 1} \defeq j$ and $a_0 \defeq i$ the \textit{unique} $j \rightarrow i$ path in $\gcg$, we will expand the representation of equation (\ref{eqn:parent_expansion}) backwards along this path:

@@

\begin{corollary}
  \label{cor:gc_implies_pwgc}
  If $\gcg$ is a strongly causal DAG then $j \gc i \implies j \pwgc i$.

  \hl{This must be true in general, should include this shortly after the definition ow PWGC}
\end{corollary}

@@

@@

We are interested in understanding under what conditions a causality
graph can be correctly recovered by pairwise testing alone.  It is not
enough that the graph simply be sparse, as the following example shows.

  This follows immediately from the fact (lemma \ref{prop:pwgc_anc})
  that $j \in \anc{i} \implies j \pwgc i$.
\end{example}

@@

\begin{lemma}
  If $\gcg$ is a strongly causal DAG then $i \pwgc j$ and $j \in \anc{i}$ are
  \textit{alternatives}, that is $i \pwgc j \implies j \notin \anc{i}$.
\end{lemma}
\begin{proof}
  Suppose that $\gcg$ is a strongly causal DAG and that we have both
  $i \pwgc j$ and $j \in \anc{i}$.  We will show that $i \in \anc{j}$,
  which produces a contradiction (since there must then be a directed cycle).

  Since $j \in \anc{i}$ there is a path $\gcgpath{j}{a}{p}{i}$.
  Moreover, since $i \pwgc j$ by lemma \ref{prop:pwgc_anc} there must
  be a common ancestor $k \in \anc{i} \cap \anc{j}$ (otherwise, we would
  have $i \in \anc{j}$, contradicting the acyclic property).  The node
  $k$ has paths to both $j,\ \gcgpath{k}{b}{q}{j}$ and $i,\ \gcgpath{k}{c}{r}{i}$.

  Now, since the graph is strongly causal, path $c$ must overlap with
  path $a$ from $j$ to $i$ otherwise there would be multiple paths
  from $k$ to $i$.  By the same reasoning with paths from $k$ to $j$,
  the path $c$ must also overlap with path $b$ from $k$ to $j$, so in
  fact path $c$ from $k$ to $i$ is
  $c = b \rightarrow j \rightarrow a$.

  \hl{TODO: We need the strengthening of prop }(\ref{prop:ancestor_properties})\hl{in order to conclude.}

\subsection{Recovering $\gcg$ via Pairwise Tests}
In this section we will show that if $\gcg$ is a strongly causal DAG, then it is possible to recover $\gcg$ via pairwise tests alone.  This is an interesting fact in and of itself, but also has some implications for applications since pairwise testing is trivially parallelizable.  In section \ref{sec:structure_learning} we will also analyze the use of pairwise testing as a heuristic for general graphs.

>>>>

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{pa(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node
\newcommand{\ancn}[2]{\mathcal{A}_{#1}(#2)}  % nth ancestors of a node
\newcommand{\gpn}[2]{gp_{#1}(#2)}  % nth generation grandparents
\newcommand{\wtalpha}[2]{\widetilde{\alpha}(#1, #2)}  % Some notation for lem:pwgc_anc
\newcommand{\dist}[2]{\mathsf{d}(#1, #2)}  % Distance between things
\newcommand{\gcgpath}[2]{#1 \rightarrow \cdots \rightarrow #2}  % A shorter path command

@@

\begin{definition}[\hl{Confounder}]
  A node $k$ will be referred to as a \textit{confounder} of nodes
  $i, j$ (neither of which are equal to $k$) if
  $k \in \anc{i} \cap \anc{j}$ and there exists a path
  $\gcgpath{k}{i}$ not containing $j$, and a path $\gcgpath{k}{i}$
  not containing $i$.
\end{definition}

@@

  Moreover, if $j \not\in \anc{i}$ then $\exists u \in \anc{i} \cap \anc{j}$ which is a confounder, that is, such that $i$ is not contained in at least one $\gcgpath{u}{j}$ path.
\end{proposition}

@@

It seems intuitive that a converse of proposition \ref{prop:ancestor_properties}
would hold, i.e. $j \in \anc{i} \implies j \pwgc i$.  Unfortunately,
this is not the case in general, as different paths through $\gcg$ can
lead to cancellation (see example \ref{ex:diamond_cancellation}).  In
fact, we do not even have $j \in \pa{i} \implies j \pwgc i$ (see
example \ref{ex:lag_cancellation}).

@@

\subsection{Strongly Causal Graphs}
The examples at the conclusion of the previous section seem rather
pathological since we have simply constructed a case where different
paths in the graph cancel exactly.  It may be possible to rid
ourselves of these pathologies for instance by defining a notion of
``robust'' pairwise causality where ``$j \pwgc _R i$'' if $j \pwgc i$
almost surely for a ``small'' random perturbation of the system
matrix.  However, we do not believe such a modification of our theory
can have useful practical implications since the difference between
paths that ``nearly cancel'' rather than cancel exactly would in
practice require inordinate amounts of data to resolve.

@@

\hl{Conjecture: A graph is strongly causal if and only if it contains
  no confounders.}

Examples of strongly causal graphs include directed trees (or
forests), and DAGs where each node has at most one parent.
Fortunately, the strongly causal property is inherited by subgraphs:

\begin{lemma}
  \label{lem:still_strongly_causal}
  If $\gcg$ is a strongly causal graph, any subgraph formed by eliminating nodes
  as well as all of the in and out edges thereof is still strongly causal.
\end{lemma}
\begin{proof}
  If $\gcg$ has at most one path between any two nodes, there can only
  be fewer paths after removing nodes from $\gcg$.
\end{proof}

We explore a number of properties of strongly causal graphs before
moving into the main result of section \ref{sec:theory} in section
\ref{sec:pairwise_algorithm}.

\begin{proposition}
  \label{prop:sc_graph_common_anc}
  In a strongly causal graph if $j \in \anc{i}$ then any
  $k \in \anc{i} \cap \anc{j}$ is not a confounder, that is,
  the unique path from $k$ to $i$ contains $j$.
\end{proposition}
\begin{proof}
  Suppose that there is a path from $k$ to $i$ which does not contain
  $j$.  In this case, there are multiple paths from $k$ to $i$ (one of
  which \textit{does} go through $j$ since $j \in \anc{i}$) which
  contradicts the assumption of strong causality.
\end{proof}

@@

\begin{proposition}
  If $\gcg$ is a strongly causal DAG then $i \pwgc j$ and $j \in \anc{i}$ are
  \textit{alternatives}, that is $i \pwgc j \implies j \notin \anc{i}$.
\end{proposition}
\begin{proof}
  Suppose that $i \pwgc j$ and $j \in \anc{i}$.  Then since $\gcg$ is
  acyclic $i \not\in \anc{j}$, and by proposition
  \ref{prop:ancestor_properties} there is some
  $k \in \anc{i}\cap\anc{j}$ which is a confounder.  However, by
  proposition \ref{prop:sc_graph_common_anc} $k$ cannot be a
  confounder, a contradiction.
\end{proof}
% This is a direct proof
  % \begin{proof}
%   Suppose that $\gcg$ is a strongly causal DAG and that we have both
%   $i \pwgc j$ and $j \in \anc{i}$, which implies that
%   $i \not \in \anc{j}$ since $\gcg$ is a DAG.  We will establish the
%   contradiction $i \in \anc{j}$.

%   Since $j \in \anc{i}$ there is a path $\gcgpath{j}{i}$.
%   Moreover, since $i \pwgc j$ by proposition \ref{prop:pwgc_anc} there
%   must be a confounding ancestor $k \in \anc{i} \cap \anc{j}$, where the
%   node $k$ has a path to $j,\ \gcgpath{k}{j}$, as well as a path
%   to $i,\ \gcgpath{k}{i}$.

%   \hl{Double check that the ancestor properties proposition implies
%     that BOTH paths do not contain the other node.}

%   Now, since the graph is strongly causal, the only possible
%   $\gcgpath{k}{i}$ path is the one obtained by concatenating the
%   $\gcgpath{k}{j}$ path with the $\gcgpath{j}{i}$ path.
%   However, this is a contradiction since $k$ is a confounder and
%   should have a $\gcgpath{k}{i}$ path which does not contain $i$.
% \end{proof}

\begin{proposition}
  \label{prop:bidirectional_edge}
  If $\gcg$ is a strongly causal DAG and $i \pwgc j$ and $j \pwgc i$,
  then $i \not\in \pa{j}$ and $j \not\in \pa{i}$.  That is, a pairwise
  bidirectional edge indicates the absense of any edge in $\gcg$.
\end{proposition}
\begin{proof}
  \hl{TODO: This is just a corollary of the previous proposition.}
% This is a direct proof
% \begin{proof}
  By way of contradiction, suppose that $i \in \pa{j}$.  We will
  consider the two possibilities allowed by proposition
  \ref{prop:ancestor_properties} for $j \pwgc i$.  Firstly
  $j \in \anc{i}$ is impossible since $\gcg$ is assumed to be acyclic.
  Secondly, if there is some confounding node
  $u \in \anc{i} \cap \anc{j}$ with a path
  $u \rightarrow \cdots \rightarrow j$ which does not contain $i$ we
  have a contradiction since there must now be multiple
  $u \rightarrow j$ paths: the aforementioned, and a path
  $u \rightarrow \cdots \rightarrow i \rightarrow \cdots \rightarrow
  j$ which \textit{does} contain $i$.  We conclude that $i \in \pa{j}$
  is impossible, and symmetrically that $j \in \pa{i}$ is as well.
% \end{proof}
\end{proof}

@@

  Firstly, we will establish a representation of $x_i(t)$ that involves $x_j(t)$.  Denote by $a_{r + 1} \rightarrow a_r \rightarrow \cdots \rightarrow a_1 \rightarrow a_0$ with $a_{r + 1} \defeq j$ and $a_0 \defeq i$ the \textit{unique} $\gcgpath{j}{i}$ path in $\gcg$, we will expand the representation of equation (\ref{eqn:parent_expansion}) backwards along this path:

@@

We immediately obtain the corollary, which we remind the reader is,
surprisingly, not true in a general graph.

\begin{corollary}
  \label{cor:gc_implies_pwgc}
  If $\gcg$ is a strongly causal DAG then $j \gc i \implies j \pwgc i$.
\end{corollary}

@@

\hl{We can possibly supply another part of the converse, as well as
  formalizing the concept described in the second part of proposition} \ref{prop:ancestor_properties}.  \hl{This also ties in nicely with proposition} \ref{prop:bidirectional_edge}

\hl{Finally, that only \textit{confounders} seem to cause problems,
  but not \textit{colliders} is a feature of the time-based nature of
  Granger-causality as opposed to Pearl's static notion of causation.}

\begin{proposition}
  If $\gcg$ is a strongly causal DAG and nodes $i, j$ have a confounder $k$, then
  $i \pwgc j$ and $j \pwgc i$.
\end{proposition}

\begin{proof}
  \hl{I'm not actually sure yet if this is true.}
\end{proof}

@@

@@

\subsection{Recovering $\gcg$ via Pairwise Tests}
\label{sec:pairwise_algorithm}
In this section we will show that if $\gcg$ is a strongly causal DAG,
then it is possible to recover $\gcg$ via pairwise tests alone.  This
is an interesting fact in and of itself, but also has some
implications for applications since pairwise testing is trivially
parallelizable.  In section \ref{sec:structure_learning} we will also
analyze the use of pairwise testing as a heuristic for general graphs.

In order to recover $\gcg$ via pairwise testing alone, it is not
enough that the graph simply be sparse, as the following example
shows.

  This follows immediately from the fact (proposition \ref{prop:pwgc_anc})
  that $j \in \anc{i} \implies j \pwgc i$.
\end{example}
